---
title: "Hospital Admissions"
author: "Onesmas Ngugi"
date: "April 4, 2018"
output: html_document
---
# INTRODUCTION

The Statewide Planning and Research Cooperative System (SPARCS) Inpatient De-identified File contains discharge level detail on patient characteristics, diagnoses, treatments, services, and charges. This data file contains basic record level detail for the discharge. The de-identified data file does not contain data that is protected health information (PHI) under HIPAA. The health information is not individually identifiable; all data elements considered identifiable have been redacted. For example, the direct identifiers regarding a date have the day and month portion of the date removed.



# Libraries and Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, cache=TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
```
```{r, eval = FALSE}
# Url for data dowload. I saved the cvs file locally to avoid dowloading each time
# I reload the script.
url <- "https://health.data.ny.gov/api/views/gnzp-ekau/rows.csv?accessType=DOWNLOAD"
download.file(url, "hosp_discharges.csv")
```
```{r}
discharges <- read_csv("hosp_discharges.csv", col_types = cols(
  `Health Service Area` = col_character(),
  `Hospital County` = col_character(),
  `Operating Certificate Number` = col_character(),
  `Facility Id` = col_integer(),
  `Facility Name` = col_character(),
  `Age Group` = col_character(),
  `Zip Code - 3 digits` = col_character(),
  Gender = col_character(),
  Race = col_character(),
  Ethnicity = col_character(),
  `Length of Stay` = col_double(),
  `Type of Admission` = col_character(),
  `Patient Disposition` = col_character(),
  `Discharge Year` = col_integer(),
  `CCS Diagnosis Code` = col_integer(),
  `CCS Diagnosis Description` = col_character(),
  `CCS Procedure Code` = col_integer(),
  `CCS Procedure Description` = col_character(),
  `APR DRG Code` = col_integer(),
  `APR DRG Description` = col_character(),
  `APR MDC Code` = col_integer(),
  `APR MDC Description` = col_character(),
  `APR Severity of Illness Code` = col_integer(),
  `APR Severity of Illness Description` = col_character(),
  `APR Risk of Mortality` = col_character(),
  `APR Medical Surgical Description` = col_character(),
  `Payment Typology 1` = col_character(),
  `Payment Typology 2` = col_character(),
  `Payment Typology 3` = col_character(),
  `Attending Provider License Number` = col_integer(),
  `Operating Provider License Number` = col_integer(),
  `Other Provider License Number` = col_integer(),
  `Birth Weight` = col_integer(),
  `Abortion Edit Indicator` = col_character(),
  `Emergency Department Indicator` = col_character(),
  `Total Charges` = col_double(),
  `Total Costs` = col_double()))
```

Let's explore the data.

```{r}
dim(discharges)
head(names(discharges))
```

'''
The dataset has `r count(discharges)` records and `r length(names(discharges))` columns.
The column names have spaces in them which makes working with them a bit cumbersome as quotes have to be used around column names everytime. We can rename the columns by replacing spaces with "_".
'''


```{r}
col_names <- str_replace_all(names(discharges), pattern = " ", replacement = "_")
names(discharges) <- col_names

# Drop Zip code column as currently unuseful.
discharges <- select(discharges, -c(7))
```

## Missing Data

Let's look at which variables have missing data. Visualizations can be a good way
to quickly show 

```{r}
sum(is.na(discharges))
```
```{r}
missing_by_column <- discharges %>% 
    is.na %>%
    as_data_frame %>%
    mutate(row_number = 1:nrow(.)) %>%
    gather(variable, is_missing, -row_number)

head(missing_by_column)
```

```{r}
ggplot(missing_by_column, aes(x = variable, y = row_number, fill = is_missing)) +
    geom_tile() +
    scale_fill_grey(name = "",
                    labels = c("Present", "Mssing")) +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +
    labs(x = "Variable",
         y = "Rows / Observations")
```


