---
title: "Hospital Admissions"
author: "Onesmas Ngugi"
date: "April 4, 2018"
output: html_document
        pdf
---
# INTRODUCTION

The Statewide Planning and Research Cooperative System (SPARCS) Inpatient De-identified File contains discharge level detail on patient characteristics, diagnoses, treatments, services, and charges. This data file contains basic record level detail for the discharge. The de-identified data file does not contain data that is protected health information (PHI) under HIPAA. The health information is not individually identifiable; all data elements considered identifiable have been redacted. For example, the direct identifiers regarding a date have the day and month portion of the date removed.

# Libraries and Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```
```{r, Import Data}
url <- "https://health.data.ny.gov/api/views/gnzp-ekau/rows.csv?accessType=DOWNLOAD"
discharges <- read_csv(url)
```



