---
title: "Hospital Admissions"
author: "Onesmas Ngugi"
date: "April 4, 2018"
output:
  html_document: default
  pdf_document: default
---
# INTRODUCTION

The Statewide Planning and Research Cooperative System (SPARCS) Inpatient De-identified File contains discharge level detail on patient characteristics, diagnoses, treatments, services, and charges. This data file contains basic record level detail for the discharge. The de-identified data file does not contain data that is protected health information (PHI) under HIPAA. The health information is not individually identifiable; all data elements considered identifiable have been redacted. For example, the direct identifiers regarding a date have the day and month portion of the date removed.

A few questions that might be interesting to help explore the data.
1. What affects hospital length of stay? Type of admission, patient disposition, 
emergency department indicator etc.
2. Are costs for similar procedures different depending in hospital area.
3. How ethnicity or race affect birth weight.
4. Costs over time.
5. Mortality rate vs severity of illness.

These are just a few ideas to help guide the exploratory analysis. 



# Libraries and Data

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)
options("max.print" = 999999)
```

```{r libraries, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(gridExtra)
sessionInfo()
```
```{r, eval = FALSE}
# Url for data dowload. I saved the cvs file locally to avoid dowloading each time
# I reload the script.
url <- "https://health.data.ny.gov/api/views/gnzp-ekau/rows.csv?accessType=DOWNLOAD"
download.file(url, "hosp_discharges.csv")
```
```{r data_import, warning=FALSE}
discharges <- read_csv("hosp_discharges.csv", na = c("", " ", "NA"), col_types = cols(
  `Health Service Area` = col_character(),
  `Hospital County` = col_character(),
  `Operating Certificate Number` = col_character(),
  `Facility Id` = col_integer(),
  `Facility Name` = col_character(),
  `Age Group` = col_character(),
  `Zip Code - 3 digits` = col_character(),
  Gender = col_character(),
  Race = col_character(),
  Ethnicity = col_character(),
  `Length of Stay` = col_double(),
  `Type of Admission` = col_character(),
  `Patient Disposition` = col_character(),
  `Discharge Year` = col_integer(),
  `CCS Diagnosis Code` = col_integer(),
  `CCS Diagnosis Description` = col_character(),
  `CCS Procedure Code` = col_integer(),
  `CCS Procedure Description` = col_character(),
  `APR DRG Code` = col_integer(),
  `APR DRG Description` = col_character(),
  `APR MDC Code` = col_integer(),
  `APR MDC Description` = col_character(),
  `APR Severity of Illness Code` = col_integer(),
  `APR Severity of Illness Description` = col_character(),
  `APR Risk of Mortality` = col_character(),
  `APR Medical Surgical Description` = col_character(),
  `Payment Typology 1` = col_character(),
  `Payment Typology 2` = col_character(),
  `Payment Typology 3` = col_character(),
  `Attending Provider License Number` = col_integer(),
  `Operating Provider License Number` = col_integer(),
  `Other Provider License Number` = col_integer(),
  `Birth Weight` = col_integer(),
  `Abortion Edit Indicator` = col_character(),
  `Emergency Department Indicator` = col_character(),
  `Total Charges` = col_double(),
  `Total Costs` = col_double()))
```

Let's explore the data.

The dataset has `r nrow(discharges)` records and `r length(discharges)` columns.
The column names have spaces in them which makes working with them a bit cumbersome as quotes have to be used around column names everytime. We can rename the columns by replacing spaces with "_".

## Data Issues 
1. Column names have spaces. Should be replaced with "_". Easier to work with.
2. I record has all entries missing. No explanation found. This record should be deleted.
3. From the data source, records relating to abortions have some information such as
i.e Facility Name, Health Service Area, Hospital County Provider Licese Number etc. 
removed for privacy purposes. We can filter these data into a separate dataframe.
4. A few columns have a lot records missing. For example, `Other_Provider_License_Number` 
column has more than 94% of records missing. This column can be dropped as there is no way to impute the missing data. Other columns with missing data are;
* `Payment_Typology_3`
* `Payment_typology_2`
* `Operating_Provider_License_Number`
* `Length_of_Stay`
5. Zip codes have last two digits removed for privacy purposes. 


### Fixes.
1. Replace spaces in column names with "_".
```{r rename_cols}
col_names <- str_replace_all(names(discharges), pattern = " ", replacement = "_")
names(discharges) <- col_names
```
2. Drop row with missing info. Identified by missing Facility_Id.
3. Remove abortion-related records.

```{r abortions}
abortion_df <- discharges %>%
    filter(Abortion_Edit_Indicator == "Y")

discharges_clean <- discharges %>%
    filter(!is.na(Facility_Id)) %>% # Row with missing info excluded.
    setdiff(abortion_df)  %>%
    select(-`Abortion_Edit_Indicator`) # Drop Abortion_Edit_Inidcator column.
```

```{r}
filter(discharges_clean, is.na(`Facility_Name`))
```

```{r}
glimpse(abortion_df)
```

We can further clean the abortion_df to drop columns with NA's but I will leave that
for later when I need to use this dataframe.


```{r}
head(discharges_clean)
```


## Explore Missing Data
Let's look at which variables have missing data. Visualizations can be a good way
to quickly communicate results to an audience. Below, I have calculated the 
percentage of missing data for each variable. Most variables have very few 
missing data.

I used the `missmap` function from the [Amelia](https://cran.r-project.org/web/packages/Amelia/vignettes/amelia.pdf) 
package to visualize missing data.

```{r na_by_col}
options(scipen=999)
sort(sapply(discharges_clean, function(x) sum(is.na(x)) / length(x) * 100), decreasing = TRUE)
```

```{r missmap}
if (!require(Amelia)) {
    install.packages(("Amelia"))
}

library(Amelia)
missmap(discharges_clean, y.labels = NULL, y.at = NULL, x.cex = 0.8) 
```

As seen above, most of the columns do not have any missing data. However, there 
are four columns that are missing a lot of data. Depending on whether the column
is useful for analysis, I can drop it or figure out how to deal with the missing
data, for example, by filling with mean or something similar.

# Univariate Analysis

In this section, I will explore individual variables to for such attributes as 
distibution, outliers etc. I will rely on histograms, boxplots and density plots 
fot visualiztions.

### Length of Stay

```{r los_hist}
p1 <- ggplot(discharges_clean, aes(Length_of_Stay)) +
    geom_histogram(binwidth = 1) 

p1_scaled <- p1 + scale_x_continuous(breaks = c(0, 10, 20, 30, 50, 75, 100, 125)) +
    theme_classic()

p2 <- ggplot(discharges_clean, aes(x = 1, y = Length_of_Stay)) +
    geom_boxplot() +
    xlab("") +
    theme_classic()

grid.arrange(p1_scaled, p2, ncol = 2)
```


```{r}
summary(discharges_clean$Length_of_Stay)
```

The plots above show that `Length_of_Stay` is right-skewed. Most of the values are 
less than 15 but there are also a lot of outliers as shown by the boxplot.

The summary shows the avergae length of stay is 5.2 days while the standard 
deviation is `r sd(discharges_clean$Length_of_Stay, na.rm = TRUE)`. 

```{r hist_func, echo = FALSE}
"Simple function for displaying histograms of numerical variables."
"... arg for additional arguments when necessary."
hist_plot <- function(data, col, ...) { 
    ggplot(data = data, aes("col")) +
        geom_histogram()
}
```

### Total Costs
```{r hist_tc}
p1 <- ggplot(discharges_clean, aes(Total_Costs)) +
    geom_histogram(bins = 50) 

p1_scaled <- p1 + 
    geom_histogram() +
    scale_x_continuous(limits = c(0, 150000)) 

grid.arrange(p1, p1_scaled, ncol = 2)
```

```{r}
summary(discharges_clean$Total_Costs)
```

The histogram on the left shows that the `Total_Cost` variable has extreme outliers
that skew the plot heavily to the right. 

On the right, I have excluded some some outliers using the scale_x_continuous 
function to make the histogram a bit more useful. However, the skeweness is still 
visible in the plot.

### Distribution of a few Discrete Variables

```{r age_group_bar}
discharges_clean %>%
    group_by(Age_Group) %>%
    tally() %>%
    ggplot(aes(Age_Group, y = (n)/sum(n))) +
        geom_bar(stat = 'identity', fill = 'red') +
        scale_y_continuous(label = scales::percent) +
        geom_text(aes(label = scales::percent(n/sum(n)),
                      y = n/sum(n)), vjust = -0.5) +
        labs(title = 'Age Group Distribution',
              y = 'Percent') +
        theme_classic()
```

```{r mixed_bars}
ggplot(discharges_clean, aes(Race, y = (..count..) / sum(..count..))) +
    geom_bar(fill = 'red') +
    scale_y_continuous(labels = scales::percent) +
    labs(title = 'Race Distibution',
         y = "Percent") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme_classic() 
```
```{r }
ggplot(discharges_clean, aes(Gender)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90))

admi_type_bar <- ggplot(discharges_clean, aes(Type_of_Admission)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90))


hsa_bar <- ggplot(discharges_clean, aes(Health_Service_Area)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90))

grid.arrange(admi_type_bar, hsa_bar, ncol = 2)
```

